Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-02-18 resex.tex
Revised introduction. Added to-do list in README file.

\begin{table*}
	\caption{\label{tab:stored_energy}%
		Stored beam energy for different past, present and future colliders. Each new machine represents a leap in stored beam energy.
	}
	\begin{ruledtabular}
		\begin{tabular}{lccccc}
			Collider& Tevatron (protons) \cite{tevatron} & LHC 2016 \cite{chamonix2017param}
			& LHC nominal \cite{lhc_design} & HL-LHC \cite{hlcdr} & FCC \cite{fcc_param_2017} \\
			\colrule
			Beam energy [TeV] & 0.98 & 6.5 & 7.0 & 7.0 & 50.0\\
			Number of bunches & 36 & 2220 & 2808 & 2748 & ? \\
			Number of particles per bunch & $2.90\times 10^{11}$ & $1.15\times 10^{11}$ & $1.15\times 10^{11}$ & $2.2\times 10^{11}$ & $1.0\times 10^{11}$\\
			Stored beam energy [MJ] & 1.6 & 265.9 & 362.2 & 678.0 & 8400 \\
		\end{tabular}
	\end{ruledtabular}
\end{table*}

\begin{figure*}
  \begin{minipage}[c]{\columnwidth}
    \includegraphics[width=\columnwidth]{passive_halo_control.png}
    \includegraphics[width=\columnwidth]{active_halo_control.png}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{\columnwidth}
  \includegraphics[width=\columnwidth]{diffusion_illustration}
  \end{minipage}
  \caption{Left: Sketch of passive halo control with a conventional
    collimation system (top) and active halo control with the addition
    of a hollow electron lens (bottom). Right: Illustration of a
    simplified model of active diffusion enhancement in the transverse
    plane. The diffusion coefficient as a function of amplitude
    (orange) is enhanced in a specific amplitude region when the
    hollow beam is turned on (from solid to dashed line). A
    corresponding reduction in beam tail population (black) is created
    (from solid to dashed line).}
  \label{fig:active_halo_control} 
\end{figure*}

Considering past, current and future high energy and intensity
colliders, each new machine has represented a considerable leap in
stored beam energy with rising values for future accelerators and
colliders (see Table~\ref{tab:stored_energy}).

Recent measurements at the LHC furthermore show that the tails of the
transverse beam distribution are overpopulated compared to a Gaussian
distribution resulting in a considerable amount of energy being stored
in the beam tails alone. In case of the LHC explicitly around 5\% of
the beam population is stored in the tails above 3.5~$\sigma$ compared
to 0.22\% in case of a Gaussian distribution leading to 19~MJ of
stored energy in the tails in case of nominal LHC parameters and 34~MJ
in case of HL-LHC~\cite{helreview_valentino}. This leads to the
conclusion that a mechanism is needed to deplete the beam tails in a
controlled manner \cite{helreview}.

The most obvious idea is to decrease the collimator gaps or scrape the
tails with a collimator type device. This approach is however not
feasible as it would generate unacceptably large loss spikes due to
the high intensity halo population in the LHC. Most promising are
methods, which increase the diffusion speed in the region of the halo
particles resulting in a smooth and continuous removal of the high
amplitude tails while leaving the core of the beam unperturbed. The
diffusing halo particles are then intercepted by the collimation
system and removed. This concept is also referred to as active halo
control compared to a passive system, which only intercepts the halo
particles without actively controlling the diffusion speed (see
Fig.~\ref{fig:active_halo_control} for illustration).

In a recent review, the need for such an active halo control system for HL-LHC has been assessed with the conclusion that it would considerably increase the margin and reduce the risk for machine protection compared to the current LHC collimation system being a passive halo control system ~\cite{helreview}. In view of the need of active halo control for HL-LHC and for future high power accelerators like HE-LHC and FCC-hh~\cite{helhcparam2011,fcc_coll_ipac2017}, different active halo control methods have been studied in recent years~\cite{helreview_bruce}, of which the hollow electron lens (HEL) is considered the superior device for the HL-LHC~\cite{helreview}. The beneficial effect of a future HEL for HL-LHC in terms of machine protection and beam collimation can not be, however, deployed at the expenses of the beam core as latter would ultimately lead to a degradation of the performance due to increased particle losses and emittance growth. In this paper, we will concentrate on the assessment in simulations and measurements of possible detrimental effets on the beam core. We will summarize possible sources of perturbations of the proton beam core focusing in particular on the case of pulsed operation and the beam experiments at the LHC conducted in this context.

This paper is structured as follows: Section~\ref{sec:hel} gives an introduction to the concept of HELs and summarizes the design parameters of the HL-LHC HELs. Section~\ref{sec:core} is dedicated to describing the sources of a residual field from the HEL in the core region. Section~\ref{sec:exp} then describes in detail the two beam experiments conducted at the LHC studying the effect of pulsed operation on the proton beam core.

\section{Hollow electron lens for HL-LHC\label{sec:hel}}
\subsection{General Overview\label{sec:hel:intro}}
Electron lenses produce DC or pulsed low energy electron beams, where the electron beam is generated in an electron gun, then guided and confined along a straight section by strong solenoids and finally dumped onto a collector. As an example the conceptual design of the HL-LHC HEL is shown in Fig.~\ref{fig:hel_layout}.
\begin{figure}[h]
	\includegraphics[width=1.0\linewidth]{hel_layout_epbeam}% Here is how to import EPS art
	\caption{\label{fig:hel_layout} Layout of HL-LHC HEL. Courtesy of CERN EN-MME group.}
\end{figure}

>>>>

\begin{table*}
  \caption{ Stored beam energy for different past, present and future
    colliders. Each new machine represents a leap in stored beam
    energy.}
  \label{tab:stored_energy}
  \begin{ruledtabular}
    \begin{tabular}{lccccc}
      Collider& Tevatron (protons) \cite{tevatron} & LHC 2016 \cite{chamonix2017param}	& LHC nominal \cite{lhc_design} & HL-LHC \cite{hlcdr} & FCC \cite{fcc_param_2017} \\
      \colrule
      Beam energy [TeV] & 0.98 & 6.5 & 7.0 & 7.0 & 50.0\\
      Number of bunches & 36 & 2220 & 2808 & 2748 & ? \\
      Number of particles per bunch & $2.90\times 10^{11}$ & $1.15\times 10^{11}$ & $1.15\times 10^{11}$ & $2.2\times 10^{11}$ & $1.0\times 10^{11}$\\
      Stored beam energy [MJ] & 1.6 & 265.9 & 362.2 & 678.0 & 8400 \\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

\begin{figure*}
  \begin{minipage}[c]{\columnwidth}
    \includegraphics[width=\columnwidth]{passive_halo_control.png}
    \includegraphics[width=\columnwidth]{active_halo_control.png}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{\columnwidth}
  \includegraphics[width=\columnwidth]{diffusion_illustration}
  \end{minipage}
  \caption{Left: Sketch of passive halo control with a conventional
    collimation system (top) and active halo control, with the
    addition of a hollow electron lens (bottom). Right: Illustration
    of a simplified model of active diffusion enhancement in the
    transverse plane. The diffusion coefficient as a function of
    amplitude (orange) is enhanced in a specific amplitude region when
    the hollow beam is turned on (from solid to dashed line). A
    corresponding reduction in beam tail population (black) is created
    (from solid to dashed line).}
  \label{fig:active_halo_control} 
\end{figure*}

Considering past, current and future high energy colliders, each new
machine has represented a considerable leap in stored beam energy
(Table~\ref{tab:stored_energy}).

Furthermore, recent measurements at the LHC show that the tails of the
transverse beam distribution are overpopulated compared to a Gaussian
distribution. This results in a considerable amount of energy being
stored in the beam tails. In particular, in the case of the LHC, about
5\% of the beam population is stored in the tails (i.e., above
3.5$\sigma$, where $\sigma$ is the Gaussian rms core beam size),
compared to 0.22\% in an ideal Gaussian distribution, leading to 19~MJ
of stored energy for nominal LHC parameters and 34~MJ in the case of
HL-LHC~\cite{helreview_valentino}. This leads to the conclusion that a
mechanism is needed to deplete the beam tails in a controlled
manner. Further information on the needs for halo control in LHC can
be found in Ref.~\cite{helreview}.

The most direct approach is to decrease the collimator gaps or to
periodically scrape the tails. However, this is not feasible, as it
would generate unacceptably large loss spikes and possibly component
damage. Most promising are methods which increase the diffusion speed
in the region of the halo particles, resulting in a smooth and
continuous removal of the high amplitude tails, while leaving the core
of the beam unperturbed. The diffusing halo particles are then
intercepted by the collimation system and removed. This concept is
also referred to as active halo control, designed to enhance a
conventional passive system, which is still needed to robustly
intercept the halo particles. An illustration of the concept is shown
in Fig.~\ref{fig:active_halo_control}.

In a recent review, the need for such an active halo control system
for HL-LHC has been assessed, with the conclusion that it would
considerably increase the operational margins and reduce the risks for
machine protection~\cite{helreview}. In view of the need of active
halo control for HL-LHC and for future high power accelerators, like
HE-LHC and FCC-hh~\cite{helhcparam2011, fcc_coll_ipac2017}, different
active halo control methods have been
studied~\cite{helreview_bruce}. The hollow electron lens (HEL) is
considered the most established, flexible and suitable technology for
the HL-LHC~\cite{helreview}.

However, the beneficial effects of an HL-LHC HEL for machine
protection and for collimation cannot come at the expense of
performance degradation due to losses or emittance growth in the beam
core. In standard electron-lens operation, a proton bunch sees the
same electron current at every revolution. It is also possible to have
different currents for different groups of bunches. Under these
conditions, the imperfections of the hollow beam have a negligible
effect. On the other hand, in order to extend the range of achievable
removal rates, pulsed operation is also being considered. In this
case, different currents can be set to act on the same bunch at each
turn. If a residual field is present at the location of the beam core,
core particles can be exposed to resonant transverse kicks and to
noise.

In this paper, we concentrate on the experimental and numerical
assessment of possible detrimental effects on the beam core of a
pulsed electron lens. Section~\ref{sec:hel} gives an introduction to
the concept of HELs and summarizes the design parameters of the HL-LHC
HELs. Section~\ref{sec:core} is dedicated to the sources of residual
fields from the HEL in the core region. Section~\ref{sec:exp}
describes the setup of simulations and experimental
conditions. Results and comparisons between simulations and
measurements are given in Section~\ref{sec:simex}, with discussion and
summary in Section~\ref{sec:sum}. Auxiliary data and plots are
collected in the Appendices.

\section{Hollow electron lens for HL-LHC}
\label{sec:hel}

\subsection{General Overview\label{sec:hel:intro}}
Electron lenses produce DC or pulsed low energy electron beams, where the electron beam is generated in an electron gun, then guided and confined along a straight section by strong solenoids and finally dumped onto a collector. As an example the conceptual design of the HL-LHC HEL is shown in Fig.~\ref{fig:hel_layout}.
\begin{figure}[h]
	\includegraphics[width=1.0\linewidth]{hel_layout_epbeam}% Here is how to import EPS art
	\caption{\label{fig:hel_layout} Layout of HL-LHC HEL. Courtesy of CERN EN-MME group.}
\end{figure}
